Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-06-04 manuscript/manuscript.tex
Update manuscript.tex

Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
%Here you would explain what problem you are tackling and briefly motivate your work.
%
%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
%
%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
%
%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

@@

In the first part, we show how to run GIST calculations for small molecules and interpret the output.
Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-state approach based on GIST. In addition, we discuss all steps in further detail at the end.

@@

Now, you can finally run the actual GIST analysis.
Center the molecule as above, to make sure that its position matches the grid parameters you decided on.
We recommend using either the PME or GPU implementation of GIST.
For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the "pme" option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the system is produced. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed through directly indexing the gist object, either by providing a single name or a list of names.
Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}

@@

This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
\subsection{Applying GIST for the Streptavidin-Biotin complex}
In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline pitfalls. We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
\subsubsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can make sure that the monomer conformations match the complex.

>>>>

Since the density is computed from an MD simulation, GIST can accurately treat confined water regions, which are challenging for implicit-solvent methods such as GBSA and self-consistent theories such as 3D-RISM \cite{Kovalenko1998-3drism}.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-states approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}.
%The 3D-RISM approach allows for a decomposition into entropic and enthalpic terms as well as a full localization of the underlying distributions to the solute structure.
%Here you would explain what problem you are tackling and briefly motivate your work.
%
%In this particular template, we have removed most of the usage examples which occur in \texttt{sample-document.tex} to provide a minimal template you can modify; however, we retain a couple of examples illustrating more unusual features of our templates/article class, such as the checklists, and information on algorithms and pseudocode.
%
%Keep in mind, as you prepare your manuscript, that you should plan for a representative image  which will be used to highlight your article on the journal website and publications. Usually, this would be one of your figures, but it must also be uploaded separately upon article submission. We give specific guidelines for this image on the journal website in the section on article submission (see \url{https://livecomsjournal.github.io/authors/policies/index.html#article-submission}).
%
%Additionally, for well-formatted manuscripts, we recommend that you let LaTeX handle figure/table placement for you as much as possible, so please avoid specifying strenuous float instructions like `[h!]` and `[H]` as much as possible.

@@

In the first part, we show how to run GIST calculations for small molecules and interpret the output.
Afterwards, we apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of ligand binding using an end-states approach based on GIST. In addition, we discuss all steps in further detail at the end.

@@

Now, you can finally run the actual GIST analysis.
Center the molecule as above, to verify that its position matches the grid parameters you decided on.
We recommend using either the PME or GPU implementation of GIST.
For the GPU version, cpptraj needs to be compiled with the \inlinecode{-cuda} flag.
While the GPU version is generally faster, the PME implementation matches the energy of the Amber MD engine more closely.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the "pme" option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above, an output data file (gist.dat) containing all calculated quantities for each voxel in the grid is produced. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.

@@

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a .loc property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with .loc works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}

@@

This tutorial discusses the visualization of OpenDX files in Section~\ref{sec:visualization}.
\subsection{Applying GIST for the Streptavidin-Biotin complex}
In this section we will apply GIST to the full complex and discuss each step of the process in more detail to outline potential pitfalls. We will afterwards introduce advanced analysis workflows which allow you to modify and better visualize the resulting data.
\subsubsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

Then split the equilibrated structure into biotin and streptavidin parts using \software{cpptraj\_remove\_mol.sh}.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.
